---
title: "Illustration of GSFA Application on Simulated Data"
author: "Yifan Zhou"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{gsfa_intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      results = "hold",
                      warning = FALSE,
                      comment = NA,
                      fig.align = "center")
```

## Environment set-up

Load the `GSFA` package:

```{r load-pkgs}
library(GSFA)
```

```{r}
set.seed(46568)
```

## Simulate a data set 

We generate a normal gene expression matrix $Y$ with 400 samples and 600 genes 
and a binary perturbation matrix $G$ with 2 types of perturbations according to:

$$G_{im} \overset{i.i.d.}{\sim} \text{Bern}(0.2), 
\phi_{ik} \overset{i.i.d.}{\sim} N(0,0.5) \Rightarrow Z = G \beta + \Phi,$$
$$F_{jk} \overset{i.i.d.}{\sim} \text{Bern}(0.1), 
U_{jk} \overset{i.i.d.}{\sim} N(0, 0.5) \Rightarrow W_{jk}=F_{jk}\cdot U_{jk},$$
$$E_{ij} \overset{i.i.d.}{\sim} N(0,1) \Rightarrow Y = ZW^T+E.$$
Gene expression $Y$ was generated from 5 factors, with each factor has ~0.1 of 
all genes with non-zero loading in it. 

The true association effects between factors and perturbations, $\beta$, are set to:
\begin{pmatrix}
1 & 0 & 0 & 0 & 0 \\
0 & 0.8 & 0 & 0 & 0
\end{pmatrix}

That is, the 1st factor is associated with perturbation 1, 
and the 2nd factor associated with perturbation 2. 

```{r sim-data-normal}
beta_true <- matrix(c(1, 0, 0, 0.8, 0, 0, 0, 0, 0, 0), ncol = 5)
sim_data <- normal_data_sim(N = 400, P = 600, K = 5, M = 2,
                            beta_true = beta_true,
                            pi_true = rep(0.1, 5),
                            psi_true = 0.5, G_prob = 0.2)
```

## Fit GSFA

Run GSFA given the normal expression data and binary perturbation matrix:

```{r fit-normal}
fit <- fit_gsfa_multivar(Y = sim_data$Y, G = sim_data$G,
                         K = 5, init.method = "svd",
                         prior_w_s = 10, prior_w_r = 0.2,
                         prior_beta_s = 5, prior_beta_r = 0.2,
                         niter = 1000, average_niter = 500, lfsr_niter = 500,
                         verbose = T, return_samples = T)
```

Estimated associations between factors and perturbations  
(note that factors are interchangeable, so their orders won't necessarily match the 
original):

```{r print-beta}
signif(fit$posterior_means$beta_pm[-nrow(fit$posterior_means$beta_pm), ],
       digits = 3)
```

Differential genes detected under perturbation 1:

```{r}
genes_detected1 <- which(fit$lfsr[, 1] < 0.05)
print(paste0(length(genes_detected1), " genes passed LFSR < 0.05."))

genes_factor1 <- which(sim_data$F[, 1] > 0)
num_olap1 <- length(intersect(genes_detected1, genes_detected1))
# Sensitivity:
sens1 <- num_olap1 / length(genes_factor1)
print(paste0("Sensitivity: ", signif(sens1, digits = 3)))
# Specificity:
fpr1 <- (length(genes_detected1) - num_olap1) / (sum(sim_data$F[, 1] == 0))
print(paste0("Specificity: ", signif(1 - fpr1, digits = 3)))
```

Differential genes detected under perturbation 2:

```{r}
genes_detected2 <- which(fit$lfsr[, 2] < 0.05)
print(paste0(length(genes_detected2), " genes passed LFSR < 0.05."))

genes_factor2 <- which(sim_data$F[, 2] > 0)
num_olap2 <- length(intersect(genes_detected2, genes_detected2))
# Sensitivity:
sens2 <- num_olap2 / length(genes_factor2)
print(paste0("Sensitivity: ", signif(sens2, digits = 3)))
# Specificity:
fpr2 <- (length(genes_detected2) - num_olap2) / (sum(sim_data$F[, 2] == 0))
print(paste0("Specificity: ", signif(1 - fpr2, digits = 3)))
```

## Session information

```{r session-info}
sessionInfo()
```
